Recalibrating Gravitational Wave Phenomenological Waveform Model

### 2024-01-11 src/tex/ms.tex
color fix and minor word changes

Aligned-spin, frequency-domain waveform models (such as IMRPhenomD) can be written as a
combination of amplitude and phase functions ($A$ and $\phi$ respectively):
\begin{align}\label{eq:}
	h(f,\theta,\Lambda) = A(f,\theta,\Lambda)e^{-i\phi(f,\theta,\Lambda^i)}\,,
\end{align}
where $f$ is the frequency, $\theta$ are the intrinsic parameters of the binary, and $\Lambda^i$ \corr{($i=1,2,\cdots,19$) is a set of 19 additional
parameters} which will be discussed below. 
The phase and amplitude functions are then split into three sections which represent the
inspiral, intermediate, and merger-ringdown (MR) parts of the waveform. 
During inspiral, $A$ and $\phi$ are known analytically from post-Newtonian (PN) theory;
IMRPhenomD uses the TaylorF2 model \citep{Buonanno:2009zt, Arun:2004hn} which is known up 
to 3.5PN order \kl{augmented with the next four higher order PN terms}.
To model the intermediate and MR regions, IMRPhenomD (and all waveforms in the IMRPhenom family)
uses a series of parameterizations\footnote{
	The parameterizations for both the amplitude and phase functions can be found in \citep{khan2016frequency}.
} 
which depend purely on $\Lambda^i$ and can be calibrated to numerical relativity (NR) simulations.
The three sections are then \textit{stitched} together using step functions.
Importantly, the parameterizations are chosen such that they can be made $\mathcal{C}^1$ continuous at the
boundary between each section.

@@

To compute the loss functions, we have to choose NR waveforms for calculating the mismatch.
Originally, 19 NR simulations were used to calibrate IMRPhenomD \citep{khan2016frequency}; nine of these 
are from the SXS catalog 
\footnote{We took the CoM-corrected waveforms with 4th-order extrapolation.} 
\citep{boyle2019sxs} and 10 BAM simulations. 
As BAM waveforms are not publicly available, we cannot use a training set identical to the original work.
Instead, we take the same nine waveforms from the SXS catalog plus two additional waveforms which closely mimic
two of the low mass ratio BAM simulations. 
The remaining BAM simulations have no close SXS counterparts, and are therefore not included in the training set.
Our main results therefore uses 11 NR waveforms for calibration which are listed in Tab.~\ref{tab:q148}.
Additional NR waveforms which are used for further calibration are listed in Tab.~\ref{tab:q1248}.

@@

With the purpose of improving downstream tasks such as parameter estimation in
mind, the more relevant metric of improvement is the distribution of improvement
in mismatch over the entire validation dataset. Figure~\ref{fig:q148} shows the
distribution of log-mismatches for the validation waveforms before and after the
optimization procedure. Here we show results using a constant PSD in our loss
function. One can see the distribution of the optimized waveform is skewed toward lower mismatches,
with the peak of the distribution being shifted by approximately an order of
magnitude. The median mismatch is reduced by 50\% (see vertical dotted lines) \kl{while the tail hardly changes}. When using
$\mathcal{L}_{\mathrm{norm}}$, we observe a less pronounced improvement with a 22.9\%
decrease in the median.

@@

The correlation we observe between the mismatches and the waveform's parameters
can also be used to design more physical ansatz. For example, we see the
recalibration process struggles to improve in the regime where $\chi_{1}$ is
positive and $\chi_{2}$ is negative. \kl{\sout{This means there is a dependency between
the two spins that is not captured by the current ansatz.}} A systematic scheme that
incorporates our calibration method with waveform design will be handy in
constructing new waveform models in the future.

\kl{We note that in comparison to the original IMRPhenomD calibration, we restricted the 
parameter space to $q\leq8$ and directly used NR waveform instead of hybridized SEOB-NR 
waveforms in training and validation purposes. These choices put the original IMRPhemonD 
result in a disadvantage, as both the parameter space and frequency range are restricted. 
We tested our result by training the model using a set of training waveforms of the same 
intrinsic parameters, but with BAM waveforms replaced with SEOBNR waveforms. We can see 
the result in Fig.~\ref{fig:q14818}, where the mismatch median changes by around $8\%$, 
This improvement mainly corresponds to the stitching problem discussed before. 
Nevertheless, it is known that IMRPhenomD has a large inaccuracy in the large mass-ratio 
regime, and is rarely used in this region of parameter space. Hence, we focus on showing 
AD with gradient descent, as a general and robust method, can be used to further optimize 
models independent of the training waveforms used.}

>>>>

Aligned-spin, frequency-domain waveform models (such as IMRPhenomD) can be written as a
combination of amplitude and phase functions ($A$ and $\phi$ respectively):
\begin{align}\label{eq:}
	h(f,\theta,\Lambda) = A(f,\theta,\Lambda)e^{-i\phi(f,\theta,\Lambda^i)}\,,
\end{align}
where $f$ is the frequency, $\theta$ are the intrinsic parameters of the binary, and $\Lambda^i$ \corr{($i=1,2,\cdots,19$) is a set of 19 additional
parameters} which will be discussed below. 
The phase and amplitude functions are then split into three sections which represent the
inspiral, intermediate, and merger-ringdown (MR) parts of the waveform. 
During inspiral, $A$ and $\phi$ are known analytically from post-Newtonian (PN) theory;
IMRPhenomD uses the TaylorF2 model \citep{Buonanno:2009zt, Arun:2004hn} which is known up 
to 3.5PN order \corr{augmented with the next four higher order PN terms}.
To model the intermediate and MR regions, IMRPhenomD (and all waveforms in the IMRPhenom family)
uses a series of parameterizations\footnote{
	The parameterizations for both the amplitude and phase functions can be found in \citep{khan2016frequency}.
} 
which depend purely on $\Lambda^i$ and can be calibrated to numerical relativity (NR) simulations.
The three sections are then \textit{stitched} together using step functions.
Importantly, the parameterizations are chosen such that they can be made $\mathcal{C}^1$ continuous at the
boundary between each section.

@@

To compute the loss functions, we have to choose NR waveforms for calculating the mismatch.
Originally, 19 NR simulations were used to calibrate IMRPhenomD \citep{khan2016frequency}; nine of these 
are from the SXS catalog 
\citep{boyle2019sxs} and 10 BAM simulations\footnote{\corr{We took the CoM-corrected waveforms with 4th-order extrapolation.}} . 
As BAM waveforms are not publicly available, we cannot use a training set identical to the original work.
Instead, we take the same nine waveforms from the SXS catalog plus two additional waveforms which closely mimic
two of the low mass ratio BAM simulations. 
The remaining BAM simulations have no close SXS counterparts, and are therefore not included in the training set.
Our main results therefore uses 11 NR waveforms for calibration which are listed in Tab.~\ref{tab:q148}.
Additional NR waveforms which are used for further calibration are listed in Tab.~\ref{tab:q1248}.

@@

With the purpose of improving downstream tasks such as parameter estimation in
mind, the more relevant metric of improvement is the distribution of improvement
in mismatch over the entire validation dataset. Figure~\ref{fig:q148} shows the
distribution of log-mismatches for the validation waveforms before and after the
optimization procedure. Here we show results using a constant PSD in our loss
function. One can see the distribution of the optimized waveform is skewed toward lower mismatches,
with the peak of the distribution being shifted by approximately an order of
magnitude. The median mismatch is reduced by 50\% (see vertical dotted lines) \corr{while the tail hardly changes}. When using
$\mathcal{L}_{\mathrm{norm}}$, we observe a less pronounced improvement with a 22.9\%
decrease in the median.

@@

The correlation we observe between the mismatches and the waveform's parameters
can also be used to design more physical ansatz. For example, we see the
recalibration process struggles to improve in the regime where $\chi_{1}$ is
positive and $\chi_{2}$ is negative. \corr{\sout{This means there is a dependency between
the two spins} \sout{that is not captured by the current ansatz.}} A systematic scheme that
incorporates our calibration method with waveform design will be handy in
constructing new waveform models in the future.

\corr{We note that the parameter space we use is restricted to $q\leq8$, while waveforms 
of $q=18$ are used in the original calibration trials. Also, we have directly used NR 
waveforms instead of hybridized SEOB-NR waveforms for training and validation purposes. 
These choices put the original IMRPhenomD result at a disadvantage, as both the parameter 
space and frequency range are restricted. We tested our result by training the model 
using a set of training waveforms of the same intrinsic parameters, but with BAM waveforms 
replaced with SEOBNR waveforms. We can see the result in Fig.~\ref{fig:q14818}, where the 
mismatch median decreases by about $8\%$, This mismatch improvement mainly concerns the 
stitching problem discussed before. Nevertheless, it is known that IMRPhenomD has a large 
inaccuracy in the large mass-ratio regime, and is rarely used in this region of parameter 
space. Hence, we focus on showing AD with gradient descent, as a general and robust method, 
can be used to further optimize GW waveform models with arbitrary sets of training 
waveforms used.}
